# Results and Discussion

`r if(gitbook_on){knitr::asis_output("Results of Phase 1 and Phase 2 assessments are summarized in Figure \\@ref(fig:map-interactive) with additional details provided in sections below.")}`

<br>

```{r map-interactive, fig.cap='Map of fish passage and habitat confirmation results', eval=gitbook_on}

##make colors for the priorities
pal <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", "No Fix"))

pal_phase1 <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", NA))

  
map <- leaflet::leaflet(height=500, width=780) |>
  leaflet::addTiles()  |> 
  # leafem::addMouseCoordinates(proj4 = 26911) |> ##can't seem to get it to render utms yet
  # leaflet::addProviderTiles(providers$"Esri.DeLorme") |> 
  leaflet::addProviderTiles("Esri.WorldTopoMap", group = "Topo") |>
  leaflet::addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") |> 
  leaflet::addPolygons(data = wshd_study_areas, color = "#F29A6E", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0,
              fillColor = "#F29A6E", label = wshd_study_areas$watershed_group_name) |>
  leaflet::addPolygons(data = wshds, color = "#0859C6", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0.25,
              fillColor = "#00DBFF",
              label = wshds$stream_crossing_id,
              popup = leafpop::popupTable(x = dplyr::select(wshds |> sf::st_set_geometry(NULL),
                                                     Site = stream_crossing_id,
                                                     elev_site:area_km),
                                          feature.id = F,
                                          row.numbers = F),
              group = "Phase 2") |> 
  leaflet::addLegend(
    position = "topright",
    colors = c("red", "yellow", "grey", "black"),
    labels = c("High", "Moderate", "Low", 'No fix'), opacity = 1,
    title = "Fish Passage Priorities") |> 
  
leaflet::addCircleMarkers(data=dplyr::filter(tab_map_phase_1, assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes"),
                 label = dplyr::filter(tab_map_phase_1, assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes") |>
                   dplyr::pull(pscis_crossing_id),
                 # label = tab_map_phase_1$pscis_crossing_id,
                 labelOptions = leaflet::labelOptions(noHide = F, textOnly = TRUE),
                 popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_1 |> sf::st_set_geometry(NULL) |> dplyr::filter(assess_type_phase1 == "Yes" | assess_type_reassessment == "Yes")),
                                                        Site = pscis_crossing_id, Priority = priority_phase1, Stream = stream_name, Road = road_name, `Habitat value`= habitat_value, `Barrier Result` = barrier_result, `Culvert data` = data_link, `Culvert photos` = photo_link, `Model data` = model_link),
                                             feature.id = F,
                                             row.numbers = F),
                 radius = 9,
                 fillColor = ~pal_phase1(priority_phase1),
                 color= "#ffffff",
                 stroke = TRUE,
                 fillOpacity = 1.0,
                 weight = 2,
                 opacity = 1.0,
                 group = "Phase 1") |> 
  leaflet::addPolylines(data=habitat_confirmation_tracks,
               opacity=0.75, color = '#e216c4',
               fillOpacity = 0.75, weight=5, group = "Phase 2") |>

leaflet::addAwesomeMarkers(
    lng = as.numeric(photo_metadata$gps_longitude),
    lat = as.numeric(photo_metadata$gps_latitude),
    popup = leafpop::popupImage(photo_metadata$url, src = "remote"),
    clusterOptions = leaflet::markerClusterOptions(),
    group = "Phase 2") |>


  leaflet::addCircleMarkers(
    data=tab_map_phase_2,
    label = tab_map_phase_2$pscis_crossing_id,
    labelOptions = leaflet::labelOptions(noHide = T, textOnly = TRUE),
    popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_2 |> sf::st_drop_geometry()),
                                           Site = pscis_crossing_id,
                                           Priority = priority,
                                           Stream = stream_name,
                                           Road = road_name,
                                           `Habitat (m)`= upstream_habitat_length_m,
                                           Comments = comments,
                                           `Culvert data` = data_link,
                                           `Culvert photos` = photo_link,
                                           `Model data` = model_link),
                                feature.id = F,
                                row.numbers = F),
    radius = 9,
    fillColor = ~pal(priority),
    color= "#ffffff",
    stroke = TRUE,
    fillOpacity = 1.0,
    weight = 2,
    opacity = 1.0,
    group = "Phase 2"
    ) |>

  leaflet::addLayersControl(
    baseGroups = c(
      "Esri.DeLorme",
      "ESRI Aerial"),
    overlayGroups = c("Phase 1", "Phase 2"),
    options = leaflet::layersControlOptions(collapsed = F)) |>
  leaflet.extras::addFullscreenControl() |>
  leaflet::addMiniMap(tiles = leaflet::providers$"Esri.NatGeoWorldMap",
             zoomLevelOffset = -6, width = 100, height = 100)

map



```


## Site Assessment Data

A summary of fish passage assessment procedures conducted through SERNbc since `r params$project_start_year` — including assessments, habitat confirmations, design, remediation, monitoring, fish sampling, eDNA sampling, and drone imagery — is provided in [Appendix - Assessment Data Summary](#app-site-data).

<br>

Orthoimagery and elevation model rasters generated since `r params$project_start_year` are catalogued in [Appendix - UAV Imagery](#app-uav).

<br>


```{r cd-rollup-body, include = FALSE}
cd_data_body <- readRDS("data/gis/climate_departure.rds")
cmp_body     <- cd_data_body$regional$cmp
cmp_pct_body <- cd_data_body$regional$cmp_pct
trn_body     <- cd_data_body$regional$trn
pick <- function(tbl, var, period = "annual") {
  tbl[tbl$variable == var & tbl$period == period, ]
}
tmean_d        <- pick(cmp_body,     "tmean")$difference
tmax_d         <- pick(cmp_body,     "tmax")$difference
tmin_d         <- pick(cmp_body,     "tmin")$difference
vpd_d          <- pick(cmp_body,     "vpd")$difference
prcp_p         <- pick(cmp_pct_body, "prcp")$difference
swe_p          <- pick(cmp_pct_body, "swe")$difference
snowfall_p     <- pick(cmp_pct_body, "snowfall")$difference
summer_swe_p   <- pick(cmp_pct_body, "swe",      "summer")$difference
spring_melt_p  <- pick(cmp_pct_body, "snowmelt", "spring")$difference
winter_melt_p  <- pick(cmp_pct_body, "snowmelt", "winter")$difference
doy50_d        <- pick(cmp_body, "snowmelt_doy_50")$difference
```

## Climate Departure

Four findings from the regional climate-departure analysis carry to fish-passage prioritisation.

**Warming is broad and significant region-wide.** All eight ecoregions of the study area warmed about `r sprintf("%.1f", tmean_d)` °C since 1951 — annual mean (+`r sprintf("%.1f", tmean_d)` °C), daytime maximum (+`r sprintf("%.1f", tmax_d)` °C), and overnight minimum (+`r sprintf("%.1f", tmin_d)` °C) all tell the same story (Mann-Kendall p < 0.001 in every ecoregion). The rate has been steady rather than accelerating — the 1981–present slope is shallower than the 1951–present slope. A west-to-east gradient layers on top, with the interior plateau ecoregions warming roughly 0.5 °C more than the eastern Rocky Mountain ecoregions — see [Appendix - Climate Departure](#app-climate-departure) Figure \@ref(fig:cd-map-tmean).

**The atmosphere is drying despite flat precipitation.** Vapour pressure deficit rose `r sprintf("%.2f", vpd_d)` hPa region-wide, significant in every ecoregion. Annual precipitation rose `r sprintf("%+.0f", prcp_p)` % regionally but the long-term trend is not statistically significant in any of the eight ecoregions; soil moisture is essentially flat because the warmer atmosphere is drinking the modest precipitation increase back through evaporation and transpiration.

**The snowpack signal is about timing, not quantity.** Total annual snowfall is essentially unchanged (`r sprintf("%+.0f", snowfall_p)` %); annual snow water equivalent is down `r sprintf("%.0f", -swe_p)` %, neither significant. The unambiguous signals are on the summer side and at the midpoint: summer SWE collapsed `r sprintf("%.0f", -summer_swe_p)` % and the snowmelt midpoint moved `r sprintf("%.0f", -doy50_d)` days earlier (p < 0.001 regionally and in every ecoregion). Counterbalancing rises in winter snowmelt (+`r sprintf("%.0f", winter_melt_p)` %) and spring snowmelt (+`r sprintf("%.0f", spring_melt_p)` %) point in the same direction but are noisier season-by-season. The freshet is arriving weeks earlier and the high-elevation snowpack that historically lingered into summer no longer does.

**Day-night asymmetry is present.** Overnight minimums warmed `r sprintf("%.1f", tmin_d)` °C while daytime maximums warmed `r sprintf("%.1f", tmax_d)` °C — overnight outpacing daytime is the textbook fingerprint of greenhouse warming [@karl_etal1993NewPerspective]. Mean diurnal range has narrowed by `r sprintf("%.1f", tmin_d - tmax_d)` °C across the record.

For barrier prioritisation these signals cut both ways. Where upstream reaches are cold-limited, added growing-degree-days can raise the intrinsic potential of habitat above the barrier and *increase* the value of restoring passage; where upstream reaches sit closer to the upper edge of the cold-water thermal niche, the same warming reduces usable habitat and lowers the payoff. The watershed-group breakdown in [Appendix - Climate Departure](#app-climate-departure) maps the eight-ecoregion signal onto the seven watershed groups via a percentage crosswalk.

## Planning — Habitat, Connectivity and Floodplain Modelling

### Habitat Modelling
Habitat modelling from `bcfishpass` including access model, linear spawning/rearing habitat model and lateral habitat 
connectivity models for watershed groups within our study area were updated for the spring of 2025 and are included
spatially in the collaborative GIS project.  A snapshot of these outputs related to each modeled and PSCIS stream 
crossing structure are also included within an `sqlite` database within this year's project reporting/code repository `r ngr::ngr_str_link_url(url_base = params$repo_url, url_resource = "blob/main/data/bcfishpass.sqlite", anchor_text = "here")`.


#### Statistical Support for `bcfishpass` Fish Habitat Modelling Updates
Initial mapping of stream discharge and temperature causal effects pathways for the future purpose of focusing aquatic restoration 
actions in areas of highest potential for positive impacts on fisheries values (ie. elimination of areas from intrinsic models 
where water temperatures are likely too cold to support fish production) are detailed in @hill_etal2024Spatialstream which is 
included as [Attachment - Water Temperature Modelling](#attach-bayes).



```{r flood-rollup-body, include = FALSE}
gpkg_fp    <- "data/gis/necr.gpkg"
aoi_fp     <- sf::st_read(gpkg_fp, layer = "aoi", quiet = TRUE)
streams_fp <- sf::st_read(gpkg_fp, layer = "streams", quiet = TRUE)
wb_fp      <- sf::st_read(gpkg_fp, layer = "waterbodies", quiet = TRUE)
fp_poly    <- sf::st_read(gpkg_fp, layer = "floodplain", quiet = TRUE)

aoi_fp     <- sf::st_make_valid(aoi_fp)
streams_fp <- suppressWarnings(sf::st_intersection(streams_fp, aoi_fp))
fp_poly    <- suppressWarnings(sf::st_intersection(fp_poly, aoi_fp))

aoi_area_km2    <- as.numeric(sum(sf::st_area(aoi_fp))) / 1e6
fp_area_ha      <- as.numeric(sum(sf::st_area(fp_poly))) / 1e4
fp_pct_aoi      <- 100 * fp_area_ha / (aoi_area_km2 * 100)

streams_in_fp    <- suppressWarnings(sf::st_intersection(streams_fp, fp_poly))
streams_km_total <- as.numeric(sum(sf::st_length(streams_fp))) / 1e3
streams_km_in_fp <- as.numeric(sum(sf::st_length(streams_in_fp))) / 1e3

wb_in_fp    <- suppressWarnings(sf::st_intersection(wb_fp, fp_poly))
lakes_n     <- sum(wb_in_fp$waterbody_type == "L", na.rm = TRUE)
wetlands_n  <- sum(wb_in_fp$waterbody_type == "W", na.rm = TRUE)
```

### Floodplain Delineation

Modelled functional floodplain in the Nechako River Watershed Group covers `r format(round(fp_area_ha), big.mark = ",")` ha — `r sprintf("%.1f", fp_pct_aoi)` % of the `r format(round(aoi_area_km2), big.mark = ",")` km² watershed group — attached to `r format(round(streams_km_in_fp), big.mark = ",")` km of chinook accessible order-3+ streams. Within that footprint sit `r lakes_n` lakes and `r format(wetlands_n, big.mark = ",")` mapped wetland polygons — the off-channel inventory that crossing remediations in these reaches stand to reconnect or leave severed, depending on whether design accounts for the floodplain context around the structure. The full metric table, watershed-scale map, and Murray Creek confluence detail map are in [Appendix - Floodplain Delineation](#app-floodplain).


## Fish Passage Assessemnts

Field assessments were conducted between `r format(min(form_pscis$date), format="%B %d, %Y")` and `r format(max(form_pscis$date), format="%B %d, %Y")` by Allan Irvine, R.P.Bio. and Lucy Schick, B.Sc. A total of `r form_pscis |> filter(assess_type_phase1 == "Yes"|assess_type_reassessment == "Yes") |> nrow()` Fish Passage Assessments were completed, including `r n_distinct(pscis_phase1$my_crossing_reference)` Phase 1 assessments and `r pscis_reassessments |> nrow()` reassessments.

<br>

Of the `r form_pscis |> filter(assess_type_phase1 == "Yes"|assess_type_reassessment == "Yes") |> nrow()` sites where fish passage assessments were completed, `r pscis_phase1 |> nrow()` were not yet inventoried in the PSCIS system.  This included `r pscis_phase1 |> filter(barrier_result == 'Passable') |> nrow()` crossings considered "passable", `r pscis_phase1 |> filter(barrier_result == 'Potential') |> nrow() |> english::as.english()` crossings considered a "potential" barrier, and `r pscis_phase1 |> filter(barrier_result == 'Barrier') |> nrow()` crossings were considered "barriers" according to threshold values based on culvert embedment, outlet drop, slope, diameter (relative to channel size) and length [@moe2011Fieldassessment].

<!-- Additionally, although all were considered fully passable, `r pscis_phase1 |> filter(barrier_result == 'Unknown') |> nrow() |> english::as.english()` crossings assessed were fords and ranked as "unknown" according to the provincial protocol. -->

<!-- <br> -->

<!-- Reassessments were completed at `r pscis_reassessments |> nrow()` sites where PSICS data required updating. -->

<br>

A summary of crossings assessed, a rough cost estimate for remediation, and a priority ranking for follow-up for Phase 1 sites is presented in Table \@ref(tab:cost-est-phase-1). Detailed data with photos are presented in `r if(gitbook_on){knitr::asis_output("[Appendix - Phase 1 Fish Passage Assessment Data and Photos]")}else(knitr::asis_output("[Attachment - Phase 1 Data and Photos](#attach-pdf-phase1-dat)"))`.


<br>



`r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`

<br>


```{r cost-est-phase-1, eval = T}

 tab_cost_est_phase1 |>
  select(`PSCIS ID`:`Cost Est ( $K)`) |> 
   fpr::fpr_kable(caption_text = paste0("Upstream habitat estimates and cost benefit analysis for Phase 1 assessments ranked as a 'barrier' or 'potential' barrier. ", sp_network_caption),
                  scroll = T)
 
```



## Habitat Confirmation Assessments

During the `r (params$project_year)` field assessments, habitat confirmation assessments were conducted at `r n_hab_conf_sites` sites within the `r wsg_names_hab_conf |> knitr::combine_words()` watershed groups. A total of approximately `r len_surveyed_km` km of stream was assessed, combining reaches surveyed upstream and downstream of each crossing.

<br>

As collaborative decision-making was ongoing at the time of reporting, site prioritization can be considered preliminary. Results are summarized in `r if(gitbook_on){knitr::asis_output("Figure \\@ref(fig:map-interactive) and ")}` Table \@ref(tab:tab-overview) with raw habitat `r if (exists("fish_data_complete")) "and fish sampling " else ""` data included in [Attachment - Data](#attach-dat). A summary of preliminary modelling results illustrating quantities of `r stringr::str_to_lower(model_species_name)` spawning and rearing habitat potentially available upstream of each crossing as estimated by measured/modelled channel width and upstream accessible stream length are presented in Figure \@ref(fig:plot-model-all). Detailed information for each site assessed with Phase 2 assessments, including maps, is presented in the per-site habitat-confirmation appendices: [Tributary to Stony Lake (PSCIS 126158)](#tributary-to-stony-lake---126158---appendix), [Tabor Creek (PSCIS 196085 and 203582)](#tabor-creek---196085-203582---appendix), [South Yuzkli Creek (PSCIS 196332)](#south-yuzkli-creek---196332---appendix), and [Tributary to Fraser River (PSCIS 203581 and 196076)](#tributary-to-fraser-river---203581-196076---appendix).

`r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`

<br>


```{r tab-overview, eval = TRUE}
table_phase2_overview <- function(dat, caption_text = '', font = font_set, scroll = TRUE){
  dat2 <- dat |>
    kable(caption = caption_text, booktabs = T, label = NA) |>
    kableExtra::kable_styling(c("condensed"),
                              full_width = T,
                              font_size = font) |>
    kableExtra::column_spec(column = c(11), width_min = '1.5in') |>
    kableExtra::column_spec(column = c(1:10), width_max = '1in')

  if(identical(scroll,TRUE)){
    dat2 <- dat2 |>
      kableExtra::scroll_box(width = "100%", height = "500px")
  }
  dat2
}

tab_overview |> 
  table_phase2_overview(caption_text = paste0("Overview of habitat confirmation sites. ", sp_rearing_caption), 
                        scroll = gitbook_on)

```

`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`


```{r phase-2-cv-assess-sum}
fpr::fpr_table_cv_summary(dat = pscis_phase2)  |>  
  fpr::fpr_kable(caption_text = 'Summary of Phase 2 fish passage reassessments.', scroll = F)
```


`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`


```{r cost-est-phase-2, eval=T}
tab_cost_est_phase2 |>
    dplyr::rename(
    `PSCIS ID` = pscis_crossing_id,
    Stream = stream_name,
    Road = road_name,
    `Barrier Result` = barrier_result,
    `Habitat value` = habitat_value,
    `Habitat Upstream (m)` = upstream_habitat_length_m,
    `Stream Width (m)` = avg_channel_width_m,
    Fix = crossing_fix_code,
    `Cost Est ( $K)` = cost_est_1000s,
    `Cost Benefit (m / $K)` = cost_net,
    `Cost Benefit (m2 / $K)` = cost_area_net
  ) |> 
  fpr::fpr_kable(caption_text = paste0("Cost benefit analysis for Phase 2 assessments. ", sp_rearing_caption),
                 scroll = F)

```


`r if(gitbook_on){knitr::asis_output("<br>")} else knitr::asis_output("\\pagebreak")`

```{r tab-habitat-summary}
# Match Location exactly rather than with stringr::str_like(). str_like() is
# case sensitive as of stringr 1.5.0, so str_like(Location, 'upstream') stopped
# matching the "Upstream" values that 0130-tables.R writes, and this table has
# been rendering empty ever since.
tab_hab_upstream <- tab_hab_summary |>
  dplyr::filter(Location == 'Upstream') |>
  dplyr::select(-Location) |>
  dplyr::rename(`PSCIS ID` = Site, `Length surveyed upstream (m)` = `Length Surveyed (m)`)

# Total the surveyed length only. The remaining columns are averages, which do
# not sum, so they are left blank in the total row rather than showing numbers
# a reader would take for totals.
tab_hab_upstream |>
  dplyr::mutate(dplyr::across(dplyr::everything(), as.character)) |>
  dplyr::bind_rows(
    tibble::tibble(
      `PSCIS ID` = "Total",
      `Length surveyed upstream (m)` = as.character(
        sum(tab_hab_upstream$`Length surveyed upstream (m)`, na.rm = TRUE))
    )
  ) |>
  fpr::fpr_kable(
    caption_text = paste0(
      "Summary of Phase 2 habitat confirmation details. Lengths are for the reach surveyed ",
      "upstream of each crossing only; downstream reaches are reported in the site appendices, ",
      "and the combined upstream and downstream total is given in the text above."),
    scroll = F)
```

<br>

```{r tab-wshd-sum-all}
fpr::fpr_table_wshd_sum() |> 
  fpr::fpr_kable(caption_text = paste0('Summary of watershed area statistics upstream of Phase 2 crossings.'),
                                 footnote_text = 'Elev P60 = Elevation at which 60% of the watershed area is above', scroll = F)

```

<br>

```{r plot-model-all-cap}

my_caption = paste0("Summary of potential rearing and spawning habitat upstream of habitat confirmation assessment sites. ", model_species_name," rearing and spawning models used for habitat estimates (total length of stream network <", rear_gradient, "% and <", spawn_gradient, "% gradient, respectively).")

```


```{r plot-model-all, fig.cap= my_caption, eval=T}


bcfp_xref_plot <- xref_bcfishpass_names |> 
  filter(!is.na(id_join) &
         !stringr::str_detect(bcfishpass, 'below') &
         !stringr::str_detect(bcfishpass, 'all') &
         !stringr::str_detect(bcfishpass, '_ha') &
         (stringr::str_detect(bcfishpass, 'rearing') | 
          stringr::str_detect(bcfishpass, 'spawning')))


bcfishpass_phase2_plot_prep <- bcfishpass |> 
  dplyr::mutate(dplyr::across(where(is.numeric), round, 1)) |>
  dplyr::filter(stream_crossing_id %in% (pscis_phase2 |> dplyr::pull(pscis_crossing_id))) |> 
  dplyr::select(stream_crossing_id, dplyr::all_of(bcfp_xref_plot$bcfishpass)) |> 
  dplyr::mutate(stream_crossing_id = as.factor(stream_crossing_id)) |> 
  tidyr::pivot_longer(cols = bt_rearing_km:bt_spawning_km) |> 
  dplyr::filter(value > 0.0 &
                !is.na(value)) |> 
  dplyr::mutate(
    name = dplyr::case_when(stringr::str_detect(name, '_rearing') ~ paste0(model_species_name, " rearing km"),
                            TRUE ~ name),
    name = dplyr::case_when(stringr::str_detect(name, '_spawning') ~ paste0(model_species_name, " spawning km"),
                            TRUE ~ name)
    # Use when more than one modelling species 
   # name = stringr::str_replace_all(name, '_rearing', ' rearing'),
   # name = stringr::str_replace_all(name, '_spawning', ' spawning')
  )
  
  
  
bcfishpass_phase2_plot_prep |> 
  ggplot2::ggplot(ggplot2::aes(x = stream_crossing_id, y = value)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~name, ncol = 2) +
  # ggdark::dark_theme_bw(base_size = 11) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggplot2::labs(x = "Site", y = "Modelled habitat (km)")


```

```{r fish-sampling-text, eval = F}

# ### Fish Sampling
# 
# Fish sampling was conducted at `r tab_fish_sites_sum |> n_distinct()` sites within 6 streams, with a total of `r tab_fish_summary |> dplyr::pull(count_fish) |> sum(na.rm = TRUE)` fish captured. Fork length, weight, and species were documented for each fish. Salmonids with fork lengths >60mm were PIT-tagged to facilitate long-term tracking of health and movement. Fork length data was used to delineate salmonids based on life stages: fry (0 to 65mm), parr (>65 to 110mm), juvenile (>110mm to 140mm), and adult (>140mm) by visually assessing the histograms presented in Figure \@ref(fig:fish-histogram). A summary of sites assessed is included in Table \@ref(tab:tab-fish-site-summary), and raw data is provided in [Attachment - Data](#attach-dat). A summary of density results is also presented in Figure \@ref(fig:plot-fish-all).
# 
# <br>
# 
# Results are presented in detail within the individual appendices in this report with the [2023 report](https://newgraphenvironment.github.io/fish_passage_peace_2023_reporting) updated with fish sampling data for PSCIS crossing 125131 on Chuchinka-Table FSR.  Documentation for each site can be accessed `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Assessment Data Summary appendix of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-site-data.html", anchor_text = "here")))}else{knitr::asis_output("by searching the site number in Table \\@ref(tab:tab-sites-cap) of [Appendix - Assessment Data Summary](#app-site-data) and clicking the _Link Report_")}`.
# 
# <br>
# 
# `r if(gitbook_on){knitr::asis_output("")} else knitr::asis_output("<br><br><br><br>")`
# 
# <br>


```


```{r tab-fish-site-summary, eval=FALSE}
tab_fish_sites_sum |> 
  fpr::fpr_kable(caption_text = 'Summary of electrofishing sites.')
```

<br>

```{r plot-fish-all, fig.cap='Boxplots of densities (fish/100m2) of fish captured downstream (ds) and upstream (us) by electrofishing.', eval=FALSE}

plot_fish_box_all <- fish_abund |> 
  dplyr::filter(
    !species_code %in% c('Mountain Whitefish', 
                         'Sucker (General)',
                         'NFC',
                         'Longnose Sucker',
                         'Sculpin (General)')
  ) |> 
  ggplot2::ggplot(ggplot2::aes(x = location, y = density_100m2)) +
  ggplot2::geom_boxplot(fill = NA) +
  ggplot2::facet_grid(life_stage ~ site, scales = "free_y", as.table = TRUE) +

  # ggplot2::facet_grid(site ~ species_code, scales = "fixed", as.table = TRUE) +
  ggplot2::theme(legend.position = "none", axis.title.x = ggplot2::element_blank()) +
  ggplot2::geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1) +
  ggplot2::ylab(expression(Density ~ (Fish/100 ~  m^2))) +
  ggplot2::theme_bw()
  # ggdark::dark_theme_bw()



plot_fish_box_all
```

<br>

```{r fish-histogram, fig.cap= 'Histograms of fish lengths by species.  Fish captured by electrofishing during habitat confirmation assessments.', eval = FALSE}

# Fish histogram -----------------------------------------------------------------------
bin_1 <- floor(min(fish_data_complete$length, na.rm = TRUE) / 5) * 5
bin_n <- ceiling(max(fish_data_complete$length, na.rm = TRUE) / 5) * 5
bins <- seq(bin_1, bin_n, by = 5)

# Check what species we have and filter out any we don't want
# fish_data_complete |> dplyr::distinct(species)

plot_fish_hist <- ggplot2::ggplot(
  fish_data_complete |> dplyr::filter(!species %in% c('Fish Unidentified Species', 'Sculpin (General)', 'NFC')),
  ggplot2::aes(x = length)
) +
  ggplot2::geom_histogram(breaks = bins, alpha = 0.75,
                          position = "identity", size = 0.75) +
  ggplot2::labs(x = "Fork Length (mm)", y = "Count (#)") +
  ggplot2::facet_wrap(~species) +
  ggplot2::theme_bw(base_size = 8) +
  ggplot2::scale_x_continuous(breaks = bins[seq(1, length(bins), by = 2)]) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

plot_fish_hist


```


### Environmental DNA (eDNA) sampling

```{r tab-edna-summary-prep, include=FALSE}
# Loads the Fraser-filtered eDNA inputs (snapshotted from the template repo via
# scripts/edna_inputs_snapshot.R) and computes the per-species summary below.

# Coerce the control flags to logical with no NAs. readr types these columns as
# logical when the CSV holds TRUE/FALSE/NA, so an is.logical() early return is
# NOT enough — the NAs survive, and `any()` over a vector containing NA returns
# NA, which surfaces in the appendix as an NA in the "Pos. control" column
# instead of a blank. Absent means "not a positive control", so NA -> FALSE.
norm_lgl <- function(x) {
  if (is.logical(x)) {
    x[is.na(x)] <- FALSE
    return(x)
  }
  out <- toupper(as.character(x)) == "TRUE"
  out[is.na(out)] <- FALSE
  out
}

# Used in main Results, the thematic appendix, and the per-site appendices —
# defined once here so every consumer downstream in the bookdown merge order can
# use it without redefinition. A local copy in an appendix WILL drift; that
# happened in the Peace report and produced inconsistent species labelling in a
# single build.
fmt_targets <- function(target, max_pos, retested) {
  if (length(target) == 0) return("—")
  parts <- paste0(
    target,
    "(", ifelse(is.na(max_pos), "?", max_pos), ")",
    ifelse(retested, "*", "")
  )
  paste(parts, collapse = ", ")
}

edna_bystargets_fraser <- readr::read_csv(
  "data/edna_unbc_results_2025_by_site_target.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(grepl(params$gis_project_name, source)) |>
  # Two samples carry the wrong crossing in the field form — see
  # edna_site_id_fix() in scripts/functions.R for the correction and why it
  # lives at the display layer rather than in the form.
  dplyr::mutate(site_id = edna_site_id_fix(site_id)) |>
  dplyr::mutate(
    control_blank_field           = norm_lgl(control_blank_field),
    control_blank_office          = norm_lgl(control_blank_office),
    control_species_present_field = norm_lgl(control_species_present_field)
  )

# Real environmental samples = exclude both blank types.
#   Office blanks are distilled-water controls filtered at the end of a session
#   at accommodation — they carry real GPS, just of a hotel rather than a stream,
#   so they are not sites. (Their `site_id` is borrowed from an unrelated site,
#   or is a date when none was handy.)
#   Field blanks are streamside but test protocol cleanliness, not site eDNA —
#   reported separately in the appendix and on the map's Controls layer.
edna_real_fraser <- edna_bystargets_fraser |>
  dplyr::filter(!control_blank_office, !control_blank_field)

n_edna_real_sites    <- dplyr::n_distinct(edna_real_fraser$site_id)
n_edna_field_blanks  <- edna_bystargets_fraser |>
  dplyr::filter(control_blank_field)  |> dplyr::distinct(site_id) |> nrow()
n_edna_office_blanks <- edna_bystargets_fraser |>
  dplyr::filter(control_blank_office) |> dplyr::distinct(site_id) |> nrow()

# UNBC assay code -> common name
edna_species_names <- c(
  BULT  = "Bull Trout",
  GRAY  = "Arctic Grayling",
  RAIN  = "Rainbow Trout",
  RAIN2 = "Rainbow Trout (confirmation primer)",
  CHIN  = "Chinook Salmon",
  COHO  = "Coho Salmon",
  # The Fraser is anadromous, so the SOCK ddPCR assay detects Sockeye Salmon.
  # Differs from the Peace report, where everything above the Peace Canyon Dam
  # is landlocked and the same assay is reported as Kokanee.
  SOCK  = "Sockeye Salmon",
  BURB  = "Burbot"
)

edna_summary_fraser <- edna_real_fraser |>
  dplyr::group_by(target) |>
  dplyr::summarise(
    sites_tested        = dplyr::n_distinct(site_id),
    sites_detected      = dplyr::n_distinct(site_id[any_clean]),
    sites_subthreshold  = dplyr::n_distinct(site_id[any_crude & !any_clean]),
    sites_no_detection  = dplyr::n_distinct(site_id[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::mutate(Species = dplyr::coalesce(edna_species_names[target], target)) |>
  dplyr::select(
    Species,
    `Code`                              = target,
    `Sites tested`                      = sites_tested,
    `Detected (≥4 droplets)`            = sites_detected,
    `Sub-threshold only (1-3 droplets)` = sites_subthreshold,
    `No detection (0 droplets)`         = sites_no_detection
  ) |>
  dplyr::arrange(dplyr::desc(`Detected (≥4 droplets)`), Species)
```

New in 2025, environmental DNA (eDNA) sampling was incorporated into the program at both previously assessed and newly added sites. A total of `r form_edna |> nrow()` collection events were undertaken across `r form_edna |> dplyr::distinct(stream_name) |> nrow()` streams, comprising `r n_edna_real_sites` real environmental samples, `r n_edna_field_blanks` field blanks, and `r n_edna_office_blanks` office blanks collected for quality assurance.

Per-species detection results across the `r n_edna_real_sites` real environmental samples are summarized in Table \@ref(tab:tab-edna-summary). Detections use the standard ddPCR call threshold of ≥4 positive droplets; sub-threshold signals (1-3 droplets) are tracked separately because they are not interpretable as species presence under ddPCR convention. Non-detections should be read with care: internal amplification controls were reported for only a minority of samples, and where they are absent a zero result cannot be distinguished from a reaction that failed or was inhibited. Detections are unaffected, since a well returning positive droplets demonstrably amplified. Site-level detail — the per-site detection table, field blanks, and retests — is provided in [Appendix - Environmental DNA Results](#app-edna), and the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_fraser_map.html", anchor_text = "interactive Fraser eDNA map")` shows toggleable per-species layers.

<br>

```{r tab-edna-summary}
edna_summary_fraser |>
  fpr::fpr_kable(
    caption_text = "Per-species summary of eDNA results across real environmental samples in the Fraser project area (excludes field and office blanks).",
    scroll = FALSE
  )
```

<br>

## Engineering Design

No new designs were commissioned in `r (params$project_year)` due to uncertainty related to forest harvesting activities and lack of funding
for the 50% costs of replacing structures.  All sites with a design for remediation of fish passage can be found `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Assessment Data Summary appendix of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-site-data.html", anchor_text = "here")))}else{knitr::asis_output("in Table \\@ref(tab:tab-sites-cap) of [Appendix - Assessment Data Summary](#app-site-data) by filtering using Design = yes")}`.




## Monitoring

In `r (params$project_year)`, baseline or follow up monitoring data was gathered through completion of an effectivness montoring form (sites where remediation has been completed) and/or through eDNA sampling and aquistion of aerial imagery at the following sites (Table \@ref(tab:tab-mon-overview):

- Effectiveness monitoring was conducted on Tabor Creek (PSCIS crossing 196085) on Willow Cale FSR, where previous remediation work backwatered the outlet, removed the outlet drop and installed baffles within the culvert. The crossing remains ranked a barrier to upstream passage.

- Effectiveness monitoring continued on Bittner Creek (PSCIS crossing 196200) on Forman Road, where the culvert was replaced with a bridge prior to the 2022 reassessment and now ranks passable [@irvine2024EffectivenessMonitoring]. Results are presented in the [Bittner Creek - 196200 - Monitoring Appendix](#bittner-2025).


```{r tab-mon-overview, eval = T}
# started but not sure if this is going to be helpful. 
form_monitoring |> 
 sf::st_drop_geometry() |> 
 dplyr::select(
    `PSCIS ID` = pscis_crossing_id,
    Stream = stream_name,
    Road = road_name,
    `Crossing type` = crossing_type,
    `Crossing subtype` = crossing_subtype,
    `Diameter or span (m)` = diameter_or_span_meters,
    `Length or width (m)` = length_or_width_meters,
    `Assessment comments` = assessment_comment
  ) |> 
   fpr::fpr_kable(caption_text = 'Summary of monitoring sites.', scroll = F)
  

```

<br>

## Collaborative GIS Environment

Fieldwork activities and background spatial layers are maintained in a collaborative QGIS and Mergin Maps environment (`r params$gis_project_name`), which field crews and partners use to view, edit and add data. The layers loaded at the time of writing are listed in [Appendix - Collaborative GIS Layers](#app-gis).
