# Results and Discussion

`r if(gitbook_on){knitr::asis_output("Results of Phase 1 and Phase 2 assessments are summarized in Figure \\@ref(fig:map-interactive) with additional details provided in sections below.")}`

<br>

```{r map-interactive, fig.cap='Map of fish passage and habitat confirmation results', eval=gitbook_on}

##make colors for the priorities
pal <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", "No Fix"))

pal_phase1 <- 
   leaflet::colorFactor(palette = c("red", "yellow", "grey", "black"), 
               levels = c("High", "Moderate", "Low", NA))

  
map <- leaflet::leaflet(height=500, width=780) |>
  leaflet::addTiles()  |> 
  # leafem::addMouseCoordinates(proj4 = 26911) |> ##can't seem to get it to render utms yet
  # leaflet::addProviderTiles(providers$"Esri.DeLorme") |> 
  leaflet::addProviderTiles("Esri.WorldTopoMap", group = "Topo") |>
  leaflet::addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") |> 
  leaflet::addPolygons(data = wshd_study_areas, color = "#F29A6E", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0,
              fillColor = "#F29A6E", label = wshd_study_areas$watershed_group_name) |>
  # `wshds` is NULL in a season with no Phase 2 habitat confirmation sites
  (\(m) if (is.null(wshds) || nrow(wshds) == 0) m else
    m |> leaflet::addPolygons(data = wshds, color = "#0859C6", weight = 1, smoothFactor = 0.5,
              opacity = 1.0, fillOpacity = 0.25,
              fillColor = "#00DBFF",
              label = wshds$stream_crossing_id,
              popup = leafpop::popupTable(x = dplyr::select(wshds |> sf::st_set_geometry(NULL),
                                                     Site = stream_crossing_id,
                                                     elev_site:area_km),
                                          feature.id = F,
                                          row.numbers = F),
              group = "Phase 2"))() |>
  leaflet::addLegend(
    position = "topright",
    colors = c("red", "yellow", "grey", "black"),
    labels = c("High", "Moderate", "Low", 'No fix'), opacity = 1,
    title = "Fish Passage Priorities") |> 
  
  # `assess_type_*` is recorded as "yes" in some regions and "Yes" in others, so
  # compare case-insensitively. The layer is skipped outright when nothing matches -
  # leafpop::popupTable() indexes mat[1, 1] and fails on a zero-row table.
  (\(m) if (nrow(dplyr::filter(tab_map_phase_1, tolower(assess_type_phase1) == "yes" | tolower(assess_type_reassessment) == "yes")) == 0) m else
    m |>
  leaflet::addCircleMarkers(data=dplyr::filter(tab_map_phase_1, tolower(assess_type_phase1) == "yes" | tolower(assess_type_reassessment) == "yes"),
                   label = dplyr::filter(tab_map_phase_1, tolower(assess_type_phase1) == "yes" | tolower(assess_type_reassessment) == "yes") |>
                     dplyr::pull(pscis_crossing_id),
                   # label = tab_map_phase_1$pscis_crossing_id,
                   labelOptions = leaflet::labelOptions(noHide = F, textOnly = TRUE),
                   popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_1 |> sf::st_set_geometry(NULL) |> dplyr::filter(tolower(assess_type_phase1) == "yes" | tolower(assess_type_reassessment) == "yes")),
                                                          Site = pscis_crossing_id, Priority = priority_phase1, Stream = stream_name, Road = road_name, `Habitat value`= habitat_value, `Barrier Result` = barrier_result, `Culvert data` = data_link, `Culvert photos` = photo_link, `Model data` = model_link),
                                               feature.id = F,
                                               row.numbers = F),
                   radius = 9,
                   fillColor = ~pal_phase1(priority_phase1),
                   color= "#ffffff",
                   stroke = TRUE,
                   fillOpacity = 1.0,
                   weight = 2,
                   opacity = 1.0,
                   group = "Phase 1"))() |> 
  # survey tracks are absent until the season's GPS tracks are cleaned and tagged
  (\(m) if (is.null(habitat_confirmation_tracks) || nrow(habitat_confirmation_tracks) == 0) m else
    m |> leaflet::addPolylines(data=habitat_confirmation_tracks,
               opacity=0.75, color = '#e216c4',
               fillOpacity = 0.75, weight=5, group = "Phase 2"))() |>

  # `photo_metadata` is NULL when no photos are tagged for publication
  (\(m) if (is.null(photo_metadata) || nrow(photo_metadata) == 0) m else
    m |> leaflet::addAwesomeMarkers(
    lng = as.numeric(photo_metadata$gps_longitude),
    lat = as.numeric(photo_metadata$gps_latitude),
    popup = leafpop::popupImage(photo_metadata$url, src = "remote"),
    clusterOptions = leaflet::markerClusterOptions(),
    group = "Phase 2"))() |>


  # `tab_map_phase_2` is NULL in a season with no Phase 2 habitat confirmation sites
  (\(m) if (is.null(tab_map_phase_2) || nrow(tab_map_phase_2) == 0) m else
    m |>
    leaflet::addCircleMarkers(
      data=tab_map_phase_2,
      label = tab_map_phase_2$pscis_crossing_id,
      labelOptions = leaflet::labelOptions(noHide = T, textOnly = TRUE),
      popup = leafpop::popupTable(x = dplyr::select((tab_map_phase_2 |> sf::st_drop_geometry()),
                                             Site = pscis_crossing_id,
                                             Priority = priority,
                                             Stream = stream_name,
                                             Road = road_name,
                                             `Habitat (m)`= upstream_habitat_length_m,
                                             Comments = comments,
                                             `Culvert data` = data_link,
                                             `Culvert photos` = photo_link,
                                             `Model data` = model_link),
                                  feature.id = F,
                                  row.numbers = F),
      radius = 9,
      fillColor = ~pal(priority),
      color= "#ffffff",
      stroke = TRUE,
      fillOpacity = 1.0,
      weight = 2,
      opacity = 1.0,
      group = "Phase 2"
      )
  )() |>

  leaflet::addLayersControl(
    baseGroups = c(
      "Esri.DeLorme",
      "ESRI Aerial"),
    overlayGroups = c("Phase 1", "Phase 2"),
    options = leaflet::layersControlOptions(collapsed = F)) |>
  leaflet.extras::addFullscreenControl() |>
  leaflet::addMiniMap(tiles = leaflet::providers$"Esri.NatGeoWorldMap",
             zoomLevelOffset = -6, width = 100, height = 100)

map



```


## Site Assessment Data

A summary of fish passage assessment procedures conducted through SERNbc since `r params$project_start_year` — including assessments, habitat confirmations, design, remediation, monitoring, fish sampling, eDNA sampling, and drone imagery — is provided in [Appendix - Assessment Data Summary](#app-site-data).

<br>

Orthoimagery and elevation model rasters generated since `r params$project_start_year` are catalogued in [Appendix - UAV Imagery](#app-uav).

<br>


## Planning — Habitat and Connectivity Modelling

### Habitat Modelling
Habitat modelling from `bcfishpass` including access model, linear spawning/rearing habitat model and lateral habitat 
connectivity models for watershed groups within our study area were updated for the spring of 2025 and are included
spatially in the collaborative GIS project.  A snapshot of these outputs related to each modeled and PSCIS stream 
crossing structure are also included within an `sqlite` database within this year's project reporting/code repository `r ngr::ngr_str_link_url(url_base = params$repo_url, url_resource = "blob/main/data/bcfishpass.sqlite", anchor_text = "here")`.


#### Statistical Support for `bcfishpass` Fish Habitat Modelling Updates
Initial mapping of stream discharge and temperature causal effects pathways for the future purpose of focusing aquatic restoration 
actions in areas of highest potential for positive impacts on fisheries values (ie. elimination of areas from intrinsic models 
where water temperatures are likely too cold to support fish production) are detailed in @hill_etal2024Spatialstream which is 
included as [Attachment - Water Temperature Modelling](#attach-bayes).



## Fish Passage Assessments

Field assessments were conducted between `r format(min(form_pscis$date), format="%B %d, %Y")` and `r format(max(form_pscis$date), format="%B %d, %Y")` by Allan Irvine, R.P.Bio. The `r params$project_year` field season was directed at effectiveness monitoring and environmental DNA sampling rather than at new fish passage assessments. A total of `r form_pscis |> dplyr::filter(tolower(assess_type_phase1) == "yes" | tolower(assess_type_reassessment) == "yes") |> nrow()` crossings were assessed, all of them reassessments of structures already inventoried in PSCIS. No Phase 1 assessments were completed.

<br>

Reassessments update the provincial record where a structure has been replaced or remediated, or where the existing PSCIS entry no longer reflects conditions at the crossing. They do not add new crossings to the inventory, so they do not carry the cost estimates and priority rankings that accompany Phase 1 results. The amalgamated assessment record for the study area, spanning `r params$project_start_year` to `r params$project_year`, is presented in [Appendix - Assessment Data Summary](#app-site-data).

<br>

## Habitat Confirmation Assessments

No Phase 2 habitat confirmation assessments were conducted in `r params$project_year`. Habitat confirmations characterise the quantity and quality of habitat upstream and downstream of a crossing prioritised during Phase 1 assessment; with no Phase 1 work completed this season, none were scheduled.

<br>

Habitat confirmation results from previous years in the `r wsg_names |> knitr::combine_words()` watershed groups are carried in the amalgamated record in [Appendix - Assessment Data Summary](#app-site-data), and in the reporting for those years.

<br>
## Environmental DNA (eDNA) Sampling

```{r tab-edna-summary-prep, include=FALSE}
# Loads the Fraser-filtered eDNA inputs (snapshotted from the template repo via
# scripts/edna_inputs_snapshot.R) and computes the per-species summary below.

# Coerce the control flags to logical with no NAs. readr types these columns as
# logical when the CSV holds TRUE/FALSE/NA, so an is.logical() early return is
# NOT enough — the NAs survive, and `any()` over a vector containing NA returns
# NA, which surfaces in the appendix as an NA in the "Pos. control" column
# instead of a blank. Absent means "not a positive control", so NA -> FALSE.
norm_lgl <- function(x) {
  if (is.logical(x)) {
    x[is.na(x)] <- FALSE
    return(x)
  }
  out <- toupper(as.character(x)) == "TRUE"
  out[is.na(out)] <- FALSE
  out
}

# Used in main Results, the thematic appendix, and the per-site appendices —
# defined once here so every consumer downstream in the bookdown merge order can
# use it without redefinition. A local copy in an appendix WILL drift; that
# happened in the Peace report and produced inconsistent species labelling in a
# single build.
fmt_targets <- function(target, max_pos, retested) {
  if (length(target) == 0) return("—")
  parts <- paste0(
    target,
    "(", ifelse(is.na(max_pos), "?", max_pos), ")",
    ifelse(retested, "*", "")
  )
  paste(parts, collapse = ", ")
}

edna_bystargets <- readr::read_csv(
  "data/edna_unbc_results_2025_by_site_target.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(grepl(params$gis_project_name, source)) |>
  # Two samples carry the wrong crossing in the field form — see
  # edna_site_id_fix() in scripts/functions.R for the correction and why it
  # lives at the display layer rather than in the form.
  dplyr::mutate(site_id = edna_site_id_fix(site_id)) |>
  dplyr::mutate(
    control_blank_field           = norm_lgl(control_blank_field),
    control_blank_office          = norm_lgl(control_blank_office),
    control_species_present_field = norm_lgl(control_species_present_field)
  )

# Real environmental samples = exclude both blank types.
#   Office blanks are distilled-water controls filtered at the end of a session
#   at accommodation — they carry real GPS, just of a hotel rather than a stream,
#   so they are not sites. (Their `site_id` is borrowed from an unrelated site,
#   or is a date when none was handy.)
#   Field blanks are streamside but test protocol cleanliness, not site eDNA —
#   reported separately in the appendix and on the map's Controls layer.
edna_real <- edna_bystargets |>
  dplyr::filter(!control_blank_office, !control_blank_field)

n_edna_real_sites    <- dplyr::n_distinct(edna_real$site_id)
n_edna_field_blanks  <- edna_bystargets |>
  dplyr::filter(control_blank_field)  |> dplyr::distinct(site_id) |> nrow()
n_edna_office_blanks <- edna_bystargets |>
  dplyr::filter(control_blank_office) |> dplyr::distinct(site_id) |> nrow()

# UNBC assay code -> common name
edna_species_names <- c(
  BULT  = "Bull Trout",
  GRAY  = "Arctic Grayling",
  RAIN  = "Rainbow Trout",
  RAIN2 = "Rainbow Trout (confirmation primer)",
  CHIN  = "Chinook Salmon",
  COHO  = "Coho Salmon",
  # The Fraser is anadromous, so the SOCK ddPCR assay detects Sockeye Salmon.
  # Differs from the Peace report, where everything above the Peace Canyon Dam
  # is landlocked and the same assay is reported as Kokanee.
  SOCK  = "Sockeye Salmon",
  BURB  = "Burbot"
)

edna_summary <- edna_real |>
  dplyr::group_by(target) |>
  dplyr::summarise(
    sites_tested        = dplyr::n_distinct(site_id),
    sites_detected      = dplyr::n_distinct(site_id[any_clean]),
    # UNBC calls anything under 4 droplets negative, including 2-3 droplet
    # results that were retested and did not reach the cutoff
    sites_no_detection  = dplyr::n_distinct(site_id[!any_clean]),
    .groups = "drop"
  ) |>
  dplyr::mutate(Species = dplyr::coalesce(edna_species_names[target], target)) |>
  dplyr::select(
    Species,
    `Code`                              = target,
    `Sites tested`                      = sites_tested,
    `Detected (≥4 droplets)`            = sites_detected,
    `Not detected`                      = sites_no_detection
  ) |>
  dplyr::arrange(dplyr::desc(`Detected (≥4 droplets)`), Species)
```

New in 2025, environmental DNA (eDNA) sampling was incorporated into the program at both previously assessed and newly added sites. A total of `r form_edna |> nrow()` collection events were undertaken across `r form_edna |> dplyr::distinct(stream_name) |> nrow()` streams, comprising `r n_edna_real_sites` real environmental samples, `r n_edna_field_blanks` field blanks, and `r n_edna_office_blanks` office blanks collected for quality assurance.

Per-species detection results across the `r n_edna_real_sites` real environmental samples are summarized in Table \@ref(tab:tab-edna-summary). Detections use the four-droplet call threshold applied by the laboratory; results below that threshold, including two- and three-droplet results that were retested without reaching the cutoff, are reported as not detected. Laboratory methods and per-sample results are in the UNBC technical report, available `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "UNBC-eDNA_Report-New_Graph-13July2026.pdf", anchor_text = "here")`. Site-level detail — the per-site detection table, field blanks, and retests — is provided in [Appendix - Environmental DNA Results](#app-edna), and the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_map.html", anchor_text = "interactive eDNA map")` shows toggleable per-species layers.

<br>

```{r tab-edna-summary}
edna_summary |>
  fpr::fpr_kable(
    caption_text = "Per-species summary of eDNA results across real environmental samples in the project area (excludes field and office blanks).",
    scroll = FALSE
  )
```

<br>

## Engineering Design

No new designs were commissioned in `r (params$project_year)` due to uncertainty related to forest harvesting activities and lack of funding
for the 50% costs of replacing structures.  All sites with a design for remediation of fish passage can be found `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("online within the Assessment Data Summary appendix of the report found ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-site-data.html", anchor_text = "here")))}else{knitr::asis_output("in Table \\@ref(tab:tab-sites-cap) of [Appendix - Assessment Data Summary](#app-site-data) by filtering using Design = yes")}`.




## Monitoring

In `r (params$project_year)`, baseline or follow up monitoring data was gathered through completion of an effectivness montoring form (sites where remediation has been completed) and/or through eDNA sampling and aquistion of aerial imagery at the following sites (Table \@ref(tab:tab-mon-overview):

- Effectiveness monitoring was conducted on Peacock Creek (PSCIS crossing 197962) on the Morice FSR, where the oval culvert was replaced with a bridge in 2021. Environmental DNA sampling upstream and downstream of the crossing confirmed coho, rainbow trout and bull trout in the reach. Results are presented in the [Peacock Creek - 197962 - Monitoring Appendix](#peacock-2025).

- Effectiveness monitoring was conducted on Waterfall Creek (PSCIS crossing 124421) at 11th Avenue, where the culvert was removed by the Gitksan Watershed Authorities. Riprap at the site has been buried and seeded, with good regrowth of grasses established and shrub planting planned. Results are presented in the [Waterfall Creek - 124421 - Monitoring Appendix](#waterfall-2025).


```{r tab-mon-overview, eval = T}
# started but not sure if this is going to be helpful. 
form_monitoring |> 
 sf::st_drop_geometry() |> 
 dplyr::select(
    `PSCIS ID` = pscis_crossing_id,
    Stream = stream_name,
    Road = road_name,
    `Crossing type` = crossing_type,
    `Crossing subtype` = crossing_subtype,
    `Diameter or span (m)` = diameter_or_span_meters,
    `Length or width (m)` = length_or_width_meters,
    `Assessment comments` = assessment_comment
  ) |> 
   fpr::fpr_kable(caption_text = 'Summary of monitoring sites.', scroll = F)
  

```

<br>

## Collaborative GIS Environment

Fieldwork activities and background spatial layers are maintained in a collaborative QGIS and Mergin Maps environment (`r params$gis_project_name`), which field crews and partners use to view, edit and add data. The layers loaded at the time of writing are listed in [Appendix - Collaborative GIS Layers](#app-gis).
