---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Salmon Stock Assessment Data {-#app-stock-assessment}


Fisheries and Oceans Canada stock assessment data was accessed via the [NuSEDS-New Salmon Escapement Database System](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6)
through the [Open Government Portal](https://open.canada.ca/en) with results presented `r if(gitbook_on) { knitr::asis_output("in Table \\@ref(tab:stock-assess-dat-all)") } else { knitr::asis_output(paste0("online in this appendix at ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-stock-assessment.html", anchor_text = "app-stock-assessment.html"))) }` [@fisheriesandoceanscanadaNuSEDSNewSalmon]. A brief memo on the data extraction process is available [here](http://htmlpreview.github.io/?https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/blob/main/memos/dfo_stock_assess.html).

`r if(identical(gitbook_on, FALSE)){knitr::asis_output("<br>")}`

```{r stock-assess-download, eval=FALSE}
library("ckanr")

# set up the connection to the data portal
ckanr_setup(url = "https://open.canada.ca/data/en")

nuseds <- rgovcan::govcan_search(keywords = c("Fraser NuSEDS"), records = 150, format_results = TRUE) |>  
  pull(resources) %>% 
  bind_rows()

# Load the most recent Nuseds - currently 20251103 (update Nov 2025)
url <- "https://api-proxy.edh-cde.dfo-mpo.gc.ca/catalogue/records/c48669a3-045b-400d-b730-48aafe8c5ee6/attachments/All%20Areas%20NuSEDS_20251103.zip"


# burn directly without bringing into memory
nuseds_all_20251103.csv
ckanr::ckan_fetch(url, store = "disk", path = "data/inputs_raw/nuseds_all_20251103.zip") 

# ignore the huge file
usethis::use_git_ignore("data/inputs_raw/nuseds_all_20251103.csv")

#read it in
all_area_nuseds <- readr::read_csv('data/inputs_raw/nuseds_all_20251103.csv')

form_pscis_2025 <- read.csv("~/Projects/repo/fish_passage_template_reporting/data/backup/2025/sern_fraser_2024/form_pscis_2025.csv")


# Big study area so lets filter using the stream names from `form_pscis`
stream_names <- form_pscis_2025 |> 
  dplyr::pull(stream_name) |> 
  stringr::str_trim() |> 
  # Must be upper case to match those in Nuseds
  stringr::str_to_upper() |> 
  unique()

# we also need to add in these waterbodies that we didn't explicitly sample but are that heavy hitters in the project area
major_streams <- "NECHAKO|CHILAKO|ENDAKO|FRANÇOIS LAKE|FRASER RIVER|MORKILL RIVER|WILLOW RIVER|TABOR RIVER|SALMON RIVER"

# Combine into one regex pattern
stream_names_regex <- c(stream_names, major_streams) |> 
  stringr::str_c(collapse = "|")

# Filter Nuseds to just include streams we visited
study_area_streams <- all_area_nuseds |> 
  dplyr::filter(stringr::str_detect(WATERBODY, stream_names_regex)) |> 
  # We have some duplicate streams so remove streams with SHORE in the name which are in the Sushwap area
  # also remove big and little salmon rivers becasue not in study area
  dplyr::filter(!stringr::str_detect(WATERBODY, "SHORE|BIG|LITTLE")) |>
  janitor::clean_names() |> 
  dplyr::arrange(waterbody)

check_names <- study_area_streams |> 
  dplyr::distinct(waterbody)

## Burn to csv. Updated name so it's clear this contains only the study area streams.
study_area_streams |> 
  readr::write_csv('data/inputs_raw/study_area_NuSEDS.csv')


```



```{r stock-assess-dat-all, results="asis", echo=FALSE, eval= gitbook_on}

my_caption <- "Fisheries and Oceans Canada stock assessment data for select watershed groups in the Fraser watershed.  Data can be filtered to specific waterbodies, columns can be dragged to reorder and filtered/reordered table can be exported to csv/excel."

# not the results="asis", echo=FALSE in the chunk header
my_tab_caption()

dfo_sad_tr <- readr::read_csv('data/inputs_raw/study_area_NuSEDS.csv') |> 
  # filtering by total_return_to_river doesn't lose any data so lets keep it for this case (it removed data after 19)
  # dplyr::filter(!is.na(total_return_to_river)) %>%
  dplyr::arrange(species, analysis_yr) |> 
  dplyr::select(waterbody,
                species, 
                analysis_yr, 
                total_return_to_river, 
                start_spawn_dt_from,
                peak_spawn_dt_from,
                end_spawn_dt_from, 
                accuracy,	
                precision,	
                index_yn,	
                # no data here
                # reliability,	
                estimate_stage,	
                estimate_classification,	
                no_inspections_used,	
                estimate_method) |> 
  janitor::clean_names(case = "title") 

dfo_sad_tr |> 
  my_dt_table()

```

<br>
