---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Salmon Stock Assessment Data {-#app-stock-assessment}

```{r stock-assess-pending, results="asis", echo=FALSE, eval = !file.exists("data/inputs_raw/study_area_NuSEDS.csv")}
knitr::asis_output(paste0(
  "The stock assessment extract has not yet been regenerated for this region. ",
  "The previous extract covered a different study area and was removed rather than ",
  "published here. See the repository issues for the regeneration task.\n\n"))
```



Fisheries and Oceans Canada stock assessment data was accessed via the [NuSEDS-New Salmon Escapement Database System](https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6)
through the [Open Government Portal](https://open.canada.ca/en) with results presented `r if(gitbook_on) { knitr::asis_output("in Table \\@ref(tab:stock-assess-dat-all)") } else { knitr::asis_output(paste0("online in this appendix at ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-stock-assessment.html", anchor_text = "app-stock-assessment.html"))) }` [@fisheriesandoceanscanadaNuSEDSNewSalmon]. A brief memo on the data extraction process is available [here](http://htmlpreview.github.io/?https://github.com/NewGraphEnvironment/restoration_wedzin_kwa_2024/blob/main/memos/dfo_stock_assess.html).

`r if(identical(gitbook_on, FALSE)){knitr::asis_output("<br>")}`

```{r stock-assess-download, eval=FALSE}
# Run by hand to regenerate `data/inputs_raw/study_area_NuSEDS.csv`. Needs network.
#
# The snapshot URL is resolved from the catalogue rather than hardcoded. DFO
# republishes NuSEDS under a new date-stamped filename and the previous one 404s -
# the hardcoded `All Areas NuSEDS_20251103.zip` here was already dead by
# 2026-08, which is why the table stopped rendering and the prose was left
# pointing at a `Table ??`.

pkg_id <- "c48669a3-045b-400d-b730-48aafe8c5ee6"

res <- jsonlite::fromJSON(
  paste0("https://open.canada.ca/data/en/api/3/action/package_show?id=", pkg_id)
)$result$resources

# the all-areas zip, whatever it is called this month
url <- res$url[grepl("^All Areas NuSEDS.*\\.zip$", res$name)][1]

# ~10 MB zipped, ~230 MB unzipped, so keep it out of the repo entirely
tmp_zip <- fs::file_temp(ext = "zip")
tmp_dir <- fs::path_temp("nuseds")
download.file(url, tmp_zip, mode = "wb")
utils::unzip(tmp_zip, exdir = tmp_dir)

all_area_nuseds <- readr::read_csv(fs::dir_ls(tmp_dir, glob = "*.csv")[1])

# repo-relative and params-driven; this pointed at another repo and another region
form_pscis_2025 <- read.csv(fs::path("data/backup", params$project_year, params$gis_project_name, paste0("form_pscis_", params$project_year, ".csv")))


# Big study area so lets filter using the stream names from `form_pscis`
stream_names <- form_pscis_2025 |>
  dplyr::pull(stream_name) |>
  stringr::str_trim() |>
  # Must be upper case to match those in Nuseds
  stringr::str_to_upper() |>
  unique()

# we also need to add in these waterbodies that we didn't explicitly sample but are that heavy hitters in the project area
major_streams <- "BULKLEY|MORICE|ZYMOETZ|KISPIOX|KITSUMKALUM|SKEENA RIVER|SUSKWA|TELKWA|BABINE"

# Combine into one regex pattern
stream_names_regex <- c(stream_names, major_streams) |>
  stringr::str_c(collapse = "|")

# Filter Nuseds to just include streams we visited
study_area_streams <- all_area_nuseds |>
  # Constrain to DFO Area 4, the Skeena, BEFORE matching on name. Stream names
  # are not unique across the coast: WATERFALL CREEK, one of this season's two
  # reassessment sites, also names a Haida Gwaii stream in area 2E, which put 224
  # rows of another region's escapement in the table. Filtering by area is the
  # discriminator; the previous name-exclusion list (`SHORE|BIG|LITTLE`) was
  # tuned to the Fraser and does nothing here.
  dplyr::filter(stringr::str_detect(AREA, "^4")) |>
  dplyr::filter(stringr::str_detect(WATERBODY, stream_names_regex)) |>
  janitor::clean_names() |>
  dplyr::arrange(waterbody)

# worth eyeballing before the write - confirms nothing out of area crept in
check_names <- study_area_streams |>
  dplyr::distinct(area, waterbody)

## Burn to csv. Updated name so it's clear this contains only the study area streams.
study_area_streams |>
  readr::write_csv('data/inputs_raw/study_area_NuSEDS.csv')


```



```{r stock-assess-dat-all, results="asis", echo=FALSE, eval = gitbook_on && file.exists("data/inputs_raw/study_area_NuSEDS.csv")}

my_caption <- "Fisheries and Oceans Canada stock assessment data for select watershed groups in the Skeena watershed.  Data can be filtered to specific waterbodies, columns can be dragged to reorder and filtered/reordered table can be exported to csv/excel."

# not the results="asis", echo=FALSE in the chunk header
my_tab_caption()

dfo_sad_tr <- readr::read_csv('data/inputs_raw/study_area_NuSEDS.csv') |> 
  # filtering by total_return_to_river doesn't lose any data so lets keep it for this case (it removed data after 19)
  # dplyr::filter(!is.na(total_return_to_river)) %>%
  dplyr::arrange(species, analysis_yr) |> 
  dplyr::select(waterbody,
                species, 
                analysis_yr, 
                total_return_to_river, 
                start_spawn_dt_from,
                peak_spawn_dt_from,
                end_spawn_dt_from, 
                accuracy,	
                precision,	
                index_yn,	
                # no data here
                # reliability,	
                estimate_stage,	
                estimate_classification,	
                no_inspections_used,	
                estimate_method) |> 
  janitor::clean_names(case = "title") 

dfo_sad_tr |> 
  my_dt_table()

```

<br>
