---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Assessment Data Summary {-#app-site-data}

Fish passage assessment procedures conducted through SERNbc since `r params$project_start_year` are amalgamated here — assessments, habitat confirmations, design, remediation, monitoring, fish sampling, eDNA sampling and drone imagery. `r if(identical(gitbook_on, FALSE)){knitr::asis_output(paste0("The detailed per-site table is interactive and is provided online within this appendix ", ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "app-site-data.html", anchor_text = "here"), "."))}else{knitr::asis_output("Table \\@ref(tab:tab-sites-sum-cap) summarizes the procedures by watershed group; Table \\@ref(tab:tab-sites-cap) gives the per-site detail.")}`

<br>

```{r ld-db}
# Read the committed snapshot rather than querying postgres, so the report
# builds from a fresh clone with no database access. The snapshot is refreshed
# with scripts/fp_inputs_snapshot.R, which pulls from the template repo and
# records the upstream commit in data/fp_inputs_snapshot_manifest.txt.
# Verified identical to `SELECT * FROM working.fp_sites_tracking` at the time of
# the swap: 757 rows, 26 within this project's watershed groups.
sites_all <- arrow::read_parquet("data/snapshots/fp_sites_tracking.parquet")
```

```{r summarize-sites}

# unique(sites_all$watershed_group_name)
# 
# # here is a list of SERN wtershed groups
# wsg <- c("Bulkley River",
#               "Zymoetz River",
#               "Kispiox River", 
#               "Kalum River", 
#               "Morice River", 
#               "Parsnip River",
#               "Carp Lake", 
#               "Crooked River")

# wsg <- c(
#               "Parsnip River",
#               "Carp Lake",
#               "Crooked River"
#               )

wsg <- c(
  "Lower Chilako River",
  "Willow River",
  "Tabor River",
  "Lower Salmon River",
  "Upper Fraser River",
  "Nechako River",
  "Morkill River",
  "Francois Lake"
)


# more straight forward is new graph only watersheds
# wsg_ng <- "Elk River"

# here is a summary with Elk watershed group removed
sites_all_summary <- sites_all |> 
  # make a flag column for uav flights
  dplyr::mutate(
    uav = dplyr::case_when(
      !is.na(link_uav1) ~ "yes",
      T ~ NA_character_
    )) |> 
  # remove the elk counts
  dplyr::filter(watershed_group %in% wsg) |>
  dplyr::group_by(watershed_group) |> 
  dplyr::summarise(
    dplyr::across(assessment:fish_sampling, ~ sum(!is.na(.x))),
    uav = sum(!is.na(uav))
  ) |> 
  sf::st_drop_geometry() |> 
  # make pretty names
  dplyr::rename_with(~ stringr::str_replace_all(., "_", " ") |> 
                       stringr::str_to_title()) |> 
  # annoying special case
  dplyr::rename(
    `Drone Imagery` = Uav) |> 
  janitor::adorn_totals()
```

```{r tab-sites-sum-cap, results="asis"}
my_caption = "Summary of fish passage assessment procedures conducted in northern British Columbia through SERNbc."
my_tab_caption()
```

```{r tab-sites-sum}
sites_all_summary |> 
  dplyr::mutate(dplyr::across(everything(), as.character)) |> 
  my_dt_table(
    page_length = 20,
    cols_freeze_left = 0
              )
```

<br>


```{r tab-sites-cap, results="asis"}
my_caption = "Details of fish passage assessment procedures conducted in northern British Columbia through SERNbc."
my_tab_caption(tip_flag = FALSE)
```

```{r tab-sites-all}
sites_all |>
  dplyr::filter(watershed_group %in% wsg) |>
  sf::st_drop_geometry() |>
  dplyr::mutate(dplyr::across(everything(), as.character)) |> 
  dplyr::relocate(watershed_group, .after = my_crossing_reference) |> 
  dplyr::select(-idx) |>
  # make pretty names
  dplyr::rename_with(~ . |>
                       stringr::str_replace_all("_", " ") |>
                       stringr::str_replace_all("repo", "Report") |>
                       stringr::str_replace_all("uav", "Drone") |>
                       stringr::str_to_title()) |> 
  # dplyr::arrange(desc(stream_crossing_id)) |> 
  
  my_dt_table(
    cols_freeze_left = 1,
    escape = FALSE
  )
```
