---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - UAV Imagery {-#app-uav}

Since `r params$project_start_year`, orthoimagery and elevation model rasters have been generated and stored as Cloud Optimized Geotiffs on a cloud service provider (AWS) with select imagery linked to in the collaborative GIS project. Additionally - a tile service has been set up to facilitate viewing and downloading of individual images, provided `r knitr::asis_output(if (gitbook_on) "in Table \\@ref(tab:tab-uav-imagery-cap)" else ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "results-and-discussion.html", anchor_text = paste0("at", params$report_url, "/results-and-discussion.html")))`.

<br>

```{r api1, eval = F}

# only needs to be run at the beginning or if we want to update

# Grab the imagery from the stac

# bc bounding box
bcbbox <-  as.numeric(
  sf::st_bbox(bcmaps::bc_bound()) |> sf::st_transform(crs = 4326)
)

# use rstac to query the collection
q <- rstac::stac("https://images.a11s.one/") |>
    rstac::stac_search(
      collections = "imagery-uav-bc-prod",
                      bbox = bcbbox
                      
                     ) |>
  rstac::post_request()

# get deets of the items
r <- q |>
  rstac::items_fetch()



```


```{r uav-clean-burn, eval = F}

# build the table to display the info
tab_uav <- tibble::tibble(url_download = purrr::map_chr(r$features, ~ purrr::pluck(.x, "assets", "image", "href"))) |> 
  dplyr::mutate(stub = stringr::str_replace_all(url_download, "https://imagery-uav-bc.s3.amazonaws.com/", "")) |> 
  tidyr::separate(
    col = stub, 
    into = c("region", "watershed_group", "year", "item", "rest"),
    sep = "/",
    extra = "drop"
  ) |> 
  dplyr::mutate(
    link_view = 
                  dplyr::case_when(
                    !tools::file_path_sans_ext(basename(url_download)) %in% c("dsm", "dtm") ~ 
                      ngr::ngr_str_link_url(
                        url_base = "https://viewer.a11s.one/?cog=",
                        url_resource = url_download, 
                        url_resource_path = FALSE,
                        # anchor_text= "URL View"
                        anchor_text= tools::file_path_sans_ext(basename(url_download))),
                    T ~ "-"),
                        link_download = ngr::ngr_str_link_url(url_base = url_download, anchor_text = url_download)
    )|> 
  dplyr::select(region, watershed_group, year, item, link_view, link_download)



# grab the imagery for this project area
project_region <- "skeena"

project_uav <- tab_uav |> 
  dplyr::filter(region == project_region)

# how many distinct items are there
# length(unique(project_uav$item))

# Burn to sqlite 
conn <- readwritesqlite::rws_connect("data/bcfishpass.sqlite")
readwritesqlite::rws_list_tables(conn)
readwritesqlite::rws_drop_table("project_uav", conn = conn)
readwritesqlite::rws_write(project_uav, exists = F, delete = TRUE,
                           conn = conn, x_name = "project_uav")
readwritesqlite::rws_disconnect(conn)

```

```{r tab-uav-imagery-cap, results="asis", eval= gitbook_on}
my_caption <- "Drone imagery download and viewer links."

my_tab_caption()
```

```{r tab-uav-imagery, eval= gitbook_on}
project_uav |>
  my_dt_table(cols_freeze_left = 2, escape = FALSE)
```

