---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Floodplain Delineation {-#app-floodplain}

```{r flood-setup, include = FALSE}
library(terra)
library(sf)
```

```{r flood-load}
gpkg    <- "data/gis/necr.gpkg"
dem_tif <- "data/gis/necr_dem.tif"
val_tif <- "data/gis/necr_valleys.tif"

present <- sf::st_layers(gpkg)$name
load_layer <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi            <- load_layer("aoi")
streams        <- load_layer("streams")
waterbodies    <- load_layer("waterbodies")
floodplain     <- load_layer("floodplain")
railways       <- load_layer("railways")
roads          <- load_layer("roads")
reserves       <- load_layer("reserves")
parks          <- load_layer("parks")
named_streams  <- load_layer("named_streams")
municipalities <- load_layer("municipalities")
dem     <- terra::rast(dem_tif)
valleys <- terra::rast(val_tif)

dem        <- terra::mask(terra::crop(dem,     terra::vect(aoi)), terra::vect(aoi))
valleys    <- terra::mask(terra::crop(valleys, terra::vect(aoi)), terra::vect(aoi))
streams     <- suppressWarnings(sf::st_intersection(streams,     aoi))
waterbodies <- suppressWarnings(sf::st_intersection(waterbodies, aoi))
floodplain  <- suppressWarnings(sf::st_intersection(floodplain,  aoi))
if (!is.null(roads))          roads          <- suppressWarnings(sf::st_intersection(roads,          aoi))
if (!is.null(railways))       railways       <- suppressWarnings(sf::st_intersection(railways,       aoi))
if (!is.null(reserves))       reserves       <- suppressWarnings(sf::st_intersection(reserves,       aoi))
if (!is.null(parks))          parks          <- suppressWarnings(sf::st_intersection(parks,          aoi))
if (!is.null(named_streams))  named_streams  <- suppressWarnings(sf::st_intersection(named_streams,  aoi))
if (!is.null(municipalities)) municipalities <- suppressWarnings(sf::st_intersection(municipalities, aoi))
```

```{r flood-rollup}
aoi_area_km2 <- as.numeric(sum(sf::st_area(aoi))) / 1e6
fp_area_ha   <- as.numeric(sum(sf::st_area(floodplain))) / 1e4
fp_pct_aoi   <- 100 * fp_area_ha / (aoi_area_km2 * 100)

streams_in_fp    <- suppressWarnings(sf::st_intersection(streams, floodplain))
streams_km_total <- as.numeric(sum(sf::st_length(streams)))    / 1e3
streams_km_in_fp <- as.numeric(sum(sf::st_length(streams_in_fp))) / 1e3

wb_in_fp    <- suppressWarnings(sf::st_intersection(waterbodies, floodplain))
lakes_n     <- sum(wb_in_fp$waterbody_type == "L", na.rm = TRUE)
wetlands_n  <- sum(wb_in_fp$waterbody_type == "W", na.rm = TRUE)
lakes_ha    <- as.numeric(sum(sf::st_area(wb_in_fp[wb_in_fp$waterbody_type == "L", ]))) / 1e4
wetlands_ha <- as.numeric(sum(sf::st_area(wb_in_fp[wb_in_fp$waterbody_type == "W", ]))) / 1e4
```

This appendix details the floodplain delineation for the Nechako River Watershed Group. The rationale for mapping floodplain extent alongside the fish passage barrier inventory is described in the report background — in brief, the off-channel habitats that floodplains sustain are often as important to fish productivity as the mainstem passage that crossing remediations restore, and infrastructure corridors can compromise floodplain function in ways that structure replacement alone does not address.

<br>

Floodplain footprints were delineated using the [`flooded`](https://newgraphenvironment.github.io/flooded/) R package, which implements the Valley Confinement Algorithm of @nagel_etal2014LandscapeScale. The algorithm combines a digital elevation model with the stream network and a modelled bankfull flood surface to distinguish valley bottoms from confined canyon reaches. Bankfull depth — the flow depth at which a stream begins to spill onto its floodplain — is estimated from upstream watershed area and mean annual precipitation using the regional regression of @hall_etal2007Predictingriver, calibrated against field-mapped floodplain sites in Pacific Northwest streams.

<br>

We mapped the **functional floodplain**: the area a river actively inundates and reworks during regular high-flow events. This sits between two other scales — the narrow active channel (the wetted perimeter at low flow) and the broader geologic valley bottom (which includes terraces untouched by water for centuries). The functional floodplain matters most for habitat planning because it is the ground where side channels form, seasonal flooding sustains wetlands, and riparian forests recruit. Inputs were the [MRDEM-30](https://open.canada.ca/data/en/dataset/18752265-bda3-498c-a4ba-9dfe68cb98da) 30 m DEM, the chinook accessible stream network (stream order ≥ 3) from [bcfishpass](https://github.com/smnorris/bcfishpass), and BC Freshwater Atlas lakes and wetlands on that network. Waterbodies were included so the valley confinement model fills cells that gradient and cost-distance masks would otherwise exclude.

<br>

```{r flood-table}
knitr::kable(
  data.frame(
    Metric = c(
      "Watershed group area (km²)",
      "Floodplain area (ha)",
      "Floodplain as % of watershed group",
      "Chinook accessible streams, order 3+ (km)",
      "Streams within floodplain (km)",
      "Lakes within floodplain (count)",
      "Lakes within floodplain (ha)",
      "Wetlands within floodplain (count)",
      "Wetlands within floodplain (ha)"
    ),
    Value = c(
      format(round(aoi_area_km2),     big.mark = ","),
      format(round(fp_area_ha),       big.mark = ","),
      sprintf("%.1f", fp_pct_aoi),
      format(round(streams_km_total), big.mark = ","),
      format(round(streams_km_in_fp), big.mark = ","),
      format(lakes_n,                 big.mark = ","),
      format(round(lakes_ha),         big.mark = ","),
      format(wetlands_n,              big.mark = ","),
      format(round(wetlands_ha),      big.mark = ",")
    ),
    stringsAsFactors = FALSE
  ),
  caption = "Nechako River Watershed Group — modelled functional floodplain and the off-channel habitat units within it.",
  col.names = c("Metric", "Value"),
  align = c("l", "r")
)
```

The Nechako River Watershed Group covers roughly
`r format(round(aoi_area_km2), big.mark = ",")` km². Modelled
functional floodplain
(Table \@ref(tab:flood-table)) covers about
`r format(round(fp_area_ha), big.mark = ",")` ha, or
`r sprintf("%.1f", fp_pct_aoi)` % of the watershed group, attached to
`r format(round(streams_km_in_fp), big.mark = ",")` km of the
`r format(round(streams_km_total), big.mark = ",")` km chinook
accessible order-3+ network. Within that footprint sit `r lakes_n`
lakes and `r format(wetlands_n, big.mark = ",")` mapped wetland
polygons — the off-channel habitat units that lateral floodplain
connectivity supports.

The watershed-wide footprint is shown in
Figure \@ref(fig:flood-map-wsg); a detail view of the Murray Creek
confluence area near Vanderhoof is shown in
Figure \@ref(fig:flood-map-detail).

```{r flood-map-wsg, fig.cap = "Nechako River Watershed Group functional floodplain (green) over hillshaded terrain. Chinook accessible order-3+ streams in blue; lakes and wetlands in light blue; parks light green; First Nations reserves light grey with diamond markers and labels; municipalities outlined in red; forest service / resource roads grey; railways black dashed; watershed boundary heavy black."}
hill <- terra::shade(
  terra::terrain(dem, "slope",  unit = "radians"),
  terra::terrain(dem, "aspect", unit = "radians")
)

terra::plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 4, 2),
            main = "Nechako River Watershed Group\nfunctional floodplain, accessible order 3+")

terra::plot(valleys, col = c(NA, "#2c7a2c"), alpha = 0.55, add = TRUE, legend = FALSE)
plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
     lwd = 0.3, add = TRUE)
plot(sf::st_geometry(streams), col = "#1f77b4", lwd = 0.4, add = TRUE)

if (!is.null(roads)) {
  fsr <- roads[roads$transport_line_type_code %in% c("RRS", "RRD", "RRN"), ]
  if (nrow(fsr) > 0L) {
    plot(sf::st_geometry(fsr), col = "#88888888", lwd = 0.25, add = TRUE)
  }
}
if (!is.null(railways)) {
  plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed", add = TRUE)
}
if (!is.null(parks)) {
  plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
       lwd = 0.7, add = TRUE)
}
if (!is.null(reserves)) {
  plot(sf::st_geometry(reserves), col = "#b2b2b288", border = "#232323",
       lwd = 0.6, add = TRUE)
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 0.9, add = TRUE)
}
if (!is.null(municipalities)) {
  plot(sf::st_geometry(municipalities), col = NA, border = "#cc0000",
       lwd = 0.8, add = TRUE)
  muni_pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(municipalities),
                                               of_largest_polygon = TRUE))
  muni_coords <- sf::st_coordinates(muni_pts)
  text(muni_coords, labels = municipalities$admin_area_name,
       cex = 0.55, col = "#cc0000", font = 2, xpd = TRUE)
}

plot(sf::st_geometry(aoi), border = "black", lwd = 1.5, add = TRUE)

if (!is.null(reserves) && "english_name" %in% names(reserves) &&
    nrow(reserves) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) text(coords[, 1] + dx * halo_x, ly,
                            labels = reserves$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  for (dy in c(-1, 1)) text(coords[, 1], ly + dy * halo_y,
                            labels = reserves$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  text(coords[, 1], ly, labels = reserves$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

```{r flood-map-detail, fig.cap = "Detail view of the Murray Creek confluence with the Nechako River. Symbology as Figure \\@ref(fig:flood-map-wsg); named streams labelled in italic blue; municipalities outlined and labelled in red."}
# Murray Creek confluence with Nechako River — Vanderhoof at bottom right.
murray <- named_streams[grepl("Murray", named_streams$gnis_name, ignore.case = TRUE), ]
if (nrow(murray) > 0L) {
  mb <- sf::st_bbox(murray)
  inset_bbox <- sf::st_bbox(c(
    xmin = unname(mb["xmin"] - (mb["xmax"] - mb["xmin"]) * 0.15),
    ymin = unname(mb["ymin"] - (mb["ymax"] - mb["ymin"]) * 0.6),
    xmax = unname(mb["xmax"] + (mb["xmax"] - mb["xmin"]) * 0.5),
    ymax = unname(mb["ymax"] + (mb["ymax"] - mb["ymin"]) * 0.15)
  ), crs = sf::st_crs(streams))
} else {
  e <- sf::st_bbox(aoi)
  cx <- (e["xmin"] + e["xmax"]) / 2
  cy <- (e["ymin"] + e["ymax"]) / 2
  span_x <- (e["xmax"] - e["xmin"]) * 0.3
  span_y <- (e["ymax"] - e["ymin"]) * 0.3
  inset_bbox <- sf::st_bbox(c(
    xmin = unname(cx - span_x), ymin = unname(cy - span_y),
    xmax = unname(cx + span_x), ymax = unname(cy + span_y)
  ), crs = sf::st_crs(streams))
}
inset_ext <- terra::ext(inset_bbox[c("xmin", "xmax", "ymin", "ymax")])

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

dem_inset            <- terra::crop(dem, inset_ext)
valleys_inset        <- terra::crop(valleys, inset_ext)
streams_inset        <- crop_sf(streams)
wb_inset             <- crop_sf(waterbodies)
aoi_inset            <- crop_sf(aoi)
roads_inset          <- crop_sf(roads)
railways_inset       <- crop_sf(railways)
reserves_inset       <- crop_sf(reserves)
parks_inset          <- crop_sf(parks)
named_streams_inset  <- crop_sf(named_streams)
municipalities_inset <- crop_sf(municipalities)

hill_inset <- terra::shade(
  terra::terrain(dem_inset, "slope",  unit = "radians"),
  terra::terrain(dem_inset, "aspect", unit = "radians")
)

terra::plot(hill_inset, col = grey(0:100 / 100), legend = FALSE, axes = FALSE,
            mar = c(2, 2, 4, 2),
            main = "Nechako River Watershed Group\nMurray Creek confluence detail")

terra::plot(valleys_inset, col = c(NA, "#2c7a2c"), alpha = 0.55,
            add = TRUE, legend = FALSE)
if (!is.null(wb_inset)) {
  plot(sf::st_geometry(wb_inset), col = "#a3cdb988", border = "#2171B5",
       lwd = 0.4, add = TRUE)
}
plot(sf::st_geometry(streams_inset), col = "#1f77b4", lwd = 0.7, add = TRUE)

if (!is.null(roads_inset)) {
  plot(sf::st_geometry(roads_inset), col = "#666666aa", lwd = 0.4, add = TRUE)
}
if (!is.null(railways_inset)) {
  plot(sf::st_geometry(railways_inset), col = "black", lwd = 1.0, lty = "dashed",
       add = TRUE)
}
if (!is.null(parks_inset)) {
  plot(sf::st_geometry(parks_inset), col = "#639b5f55", border = "#33a02c",
       lwd = 0.8, add = TRUE)
}
if (!is.null(reserves_inset)) {
  plot(sf::st_geometry(reserves_inset), col = "#b2b2b2aa", border = "#232323",
       lwd = 0.7, add = TRUE)
}
if (!is.null(municipalities_inset)) {
  plot(sf::st_geometry(municipalities_inset), col = NA, border = "#cc0000",
       lwd = 1.0, add = TRUE)
  muni_pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(municipalities_inset),
                                               of_largest_polygon = TRUE))
  muni_coords <- sf::st_coordinates(muni_pts)
  ofs_y <- par("cxy")[2] * 0.6 * 0.9
  halo_x <- par("cxy")[1] * 0.6 * 0.05
  halo_y <- par("cxy")[2] * 0.6 * 0.05
  ly <- muni_coords[, 2] - ofs_y
  for (dx in c(-1, 1)) text(muni_coords[, 1] + dx * halo_x, ly,
                            labels = municipalities_inset$admin_area_name,
                            cex = 0.6, col = "white", font = 2, xpd = TRUE)
  for (dy in c(-1, 1)) text(muni_coords[, 1], ly + dy * halo_y,
                            labels = municipalities_inset$admin_area_name,
                            cex = 0.6, col = "white", font = 2, xpd = TRUE)
  text(muni_coords[, 1], ly, labels = municipalities_inset$admin_area_name,
       cex = 0.6, col = "#cc0000", font = 2, xpd = TRUE)
}
if (!is.null(aoi_inset)) {
  plot(sf::st_geometry(aoi_inset), border = "black", lwd = 1.5, add = TRUE)
}

if (!is.null(named_streams_inset)) {
  dedup <- named_streams_inset[!duplicated(named_streams_inset$gnis_name), ]
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
  text(sf::st_coordinates(pts), labels = dedup$gnis_name,
       cex = 0.5, font = 3, col = "#0d3a6c")
}

if (!is.null(reserves_inset) && "english_name" %in% names(reserves_inset) &&
    nrow(reserves_inset) > 0L) {
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves_inset),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = 1.0, add = TRUE)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) text(coords[, 1] + dx * halo_x, ly,
                            labels = reserves_inset$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  for (dy in c(-1, 1)) text(coords[, 1], ly + dy * halo_y,
                            labels = reserves_inset$english_name,
                            cex = 0.45, col = "white", font = 2, xpd = TRUE)
  text(coords[, 1], ly, labels = reserves_inset$english_name,
       cex = 0.45, col = "#111111", font = 2, xpd = TRUE)
}
```

The watershed-wide view shows floodplain tracking the trunk Nechako River through broad valley reaches and tributary confluences. The detail view near Vanderhoof resolves how road corridors and settlement intersect the floodplain footprint at reach scale — context that informs whether crossing remediation at a given site should account for restoring lateral connectivity beyond the structure itself. We plan to extend this mapping to additional watershed groups in coming years so that floodplain context can inform barrier prioritisation and restoration scoping across the region.
