---
output: html_document
editor_options:
  chunk_output_type: console
---

# Appendix - Environmental DNA Results {-#app-edna}

This appendix carries the site-level detail behind the headline counts in the Results chapter. Field blanks are reported separately because they test the sampling protocol rather than a location. Office blanks are excluded entirely — they are distilled-water controls filtered at the end of a session at accommodation, so although they carry real GPS coordinates, those coordinates record where the filtering happened rather than a stream site. The interactive map is linked at the end.

```{r app-edna-prep, include=FALSE}
# Loads the snapshotted by_site_target CSV and prepares the per-site,
# field-blank and retest views used by the tables below.

# `norm_lgl()`, `edna_species_names` and `fmt_targets()` are defined once in
# 0400-results.Rmd's prep chunk and reach this appendix through the bookdown
# merge order (0400 < 0837). Do NOT redefine them here — a local copy in the
# Peace report drifted from the Results labelling and produced two different
# species names for the same assay in one build. Peace also keeps a second copy
# of norm_lgl() in its appendix, which is how a fix to one can miss the other.

edna_bystargets_app <- readr::read_csv(
  "data/edna_unbc_results_2025_by_site_target.csv",
  show_col_types = FALSE
) |>
  dplyr::filter(grepl(params$gis_project_name, source)) |>
  dplyr::mutate(site_id = edna_site_id_fix(site_id)) |>
  dplyr::mutate(
    control_blank_field           = norm_lgl(control_blank_field),
    control_blank_office          = norm_lgl(control_blank_office),
    control_species_present_field = norm_lgl(control_species_present_field)
  ) |>
  # Office blanks are protocol controls filtered at accommodation, not sites.
  dplyr::filter(!control_blank_office)

# --- Per-site rollup (real environmental samples only) ---------------------

edna_real_app <- edna_bystargets_app |>
  dplyr::filter(!control_blank_field)

edna_per_site <- edna_real_app |>
  dplyr::group_by(site_id, stream_name) |>
  dplyr::summarise(
    `UNBC ID`         = paste(sort(unique(unlist(strsplit(lab_ids, ", ")))), collapse = ", "),
    `Pos. control`    = ifelse(any(control_species_present_field), "yes", ""),
    `Detected`        = fmt_targets(target[any_clean],
                                    max_pos[any_clean],
                                    retested[any_clean]),
    `Sub-threshold`   = fmt_targets(target[any_crude & !any_clean],
                                    max_pos[any_crude & !any_clean],
                                    retested[any_crude & !any_clean]),
    `Not detected`    = fmt_targets(target[!any_crude],
                                    max_pos[!any_crude],
                                    retested[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::rename(`Site ID` = site_id, `Stream` = stream_name) |>
  dplyr::arrange(`Site ID`)

# --- Field blanks (separate table) -----------------------------------------

edna_field_blanks <- edna_bystargets_app |>
  dplyr::filter(control_blank_field) |>
  dplyr::group_by(site_id, stream_name) |>
  dplyr::summarise(
    `UNBC ID`            = paste(sort(unique(unlist(strsplit(lab_ids, ", ")))), collapse = ", "),
    `Detected`           = fmt_targets(target[any_clean],
                                       max_pos[any_clean],
                                       retested[any_clean]),
    `Sub-threshold`      = fmt_targets(target[any_crude & !any_clean],
                                       max_pos[any_crude & !any_clean],
                                       retested[any_crude & !any_clean]),
    `Clean (0 droplets)` = fmt_targets(target[!any_crude],
                                       max_pos[!any_crude],
                                       retested[!any_crude]),
    .groups = "drop"
  ) |>
  dplyr::rename(`Site ID` = site_id, `Stream` = stream_name) |>
  dplyr::arrange(`Site ID`)

# --- Retests (site x target combinations UNBC re-ran) ----------------------

edna_retests <- edna_bystargets_app |>
  dplyr::filter(retested) |>
  dplyr::mutate(
    Species = dplyr::coalesce(edna_species_names[target], target),
    `Result` = dplyr::case_when(
      any_clean ~ "Detected (≥4 droplets)",
      any_crude ~ "Sub-threshold (1-3 droplets, NOT a detection)",
      TRUE      ~ "No detection (0 droplets)"
    ),
    `Sample type` = dplyr::case_when(
      control_blank_field ~ "field blank",
      TRUE                ~ "environmental"
    )
  ) |>
  dplyr::select(
    `Site ID`      = site_id,
    Stream         = stream_name,
    Species,
    `Code`         = target,
    `Sample type`,
    `Runs`         = n_runs,
    `Max droplets` = max_pos,
    Result
  ) |>
  dplyr::arrange(`Site ID`, Code)
```

## Per-site detection results {-}

Format: `CODE(max_droplets)`, with `*` flagging samples that UNBC reran. Detected = ≥4 positive droplets per the standard ddPCR call threshold; Sub-threshold = 1-3 droplets, not interpretable as species presence under ddPCR convention; Not detected = 0 droplets across all runs.

```{r tab-edna-per-site}
edna_per_site |>
  fpr::fpr_kable(
    caption_text = "Site-level eDNA detection results across real environmental samples in the Fraser project area. UNBC ID is the lab sample identifier; Pos. control = yes indicates the sample was collected at a site where the target species was also confirmed by electrofishing.",
    scroll = gitbook_on
  )
```

<br>

## Field blanks {-}

Field blanks filter distilled water at the field site to test for protocol contamination during collection and processing. **Detections in blanks indicate potential cross-contamination, not species presence at the location** — these rows are presented for QA transparency only, not as eDNA detections.

```{r tab-edna-field-blanks}
edna_field_blanks |>
  fpr::fpr_kable(
    caption_text = "Field blanks collected during the 2025 eDNA program in the Fraser project area. Detections in blanks reflect protocol-side contamination, not site eDNA.",
    scroll = FALSE
  )
```

<br>

## Retests {-}

UNBC reran a subset of samples to confirm borderline results. The table below lists every site × species combination that was rerun, including reruns of both real environmental samples and field blanks.

```{r tab-edna-retests, eval = TRUE}
if (nrow(edna_retests) > 0) {
  edna_retests |>
    fpr::fpr_kable(
      caption_text = "UNBC retests for the 2025 Fraser eDNA samples. Runs is the total number of times the sample x species combination was assayed; Max droplets is the highest positive-droplet count observed across all runs.",
      scroll = gitbook_on
    )
} else {
  cat("No retests in the Fraser 2025 batch.\n")
}
```

<br>

## Interactive map {-}

The interactive Fraser eDNA map shows site-level detection results with toggleable per-species layers, an opt-in sub-threshold-signals layer (off by default), and an opt-in Controls layer for field blanks. See the `r ngr::ngr_str_link_url(url_base = params$report_url, url_resource = "edna_unbc_results_2025_fraser_map.html", anchor_text = "interactive Fraser eDNA map")`.
